\textbf{Effective model selection and implementation when applying machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-09-10 submission/manuscript.tex
changes after Pat meeting 6

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Best practices for applying machine learning to bacterial 16S
rRNA gene sequencing data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Jenna
Wiens\({^2}\), Mack Ruffin\({^3}\), Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify the microbial biomarkers and aid in diagnosis of many
chronic diseases such as inflammatory bowel disease, diabetes and
colorectal cancer. Progress has been made towards developing ML models
that predict health outcomes from bacterial abundances, but rigourous ML
models are scarce due to the flawed methods that call the validity of
developed models into question. Furthermore, the use of black box ML
models has hindered the validation of microbial biomarkers. To overcome
these challenges, we benchmarked seven different ML models that use
fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions
(n=490 patients, 261 controls and 229 cases). To show the effect of
model selection, we assessed the predictive performance,
interpretability, and computational efficiency of the following models:
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, a decision
tree, random forest, and extreme gradient boosting (XGBoost). The random
forest model was best at detecting CRC lesions with an AUROC of 0.695
but it was slow to train (83.2 h) and hard to interpret. Despite its
simplicity, L2-regularized logistic regression followed random forest in
predictive performance with an AUROC of 0.680, and it trained much
faster (12 min). In this study, we established standards for the
development of modeling pipelines for microbiome-associated ML models.
Additionally, we showed that ML models should be chosen based on
expectations of predictive performance, interpretability and available
computational resources.

@@

\subsection{Importance (needs work)}\label{importance-needs-work}

Prediction of health outcomes using ML is rapidly being adopted by human
microbiome studies. However, the developed ML models so far are
overoptimistic in terms of validity and predictive performance. Without
rigorous ML pipelines, we cannot trust ML models. Before we can speed up
progress, we need to slow down, define and implement good ML practices.
\newpage

@@

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual with high confidence
relies on a ML model that is built with rigorous methods. However, there
are common methodological problems across many of these studies that
need to be addressed as the field progresses. These include a lack of
transparency in which methods are used and how these methods are
implemented; developing and evaluating models without a separate
held-out test data; large variation between the predictive performance
on different folds of cross-validation; and large variation between
cross-validation and testing performances. Nevertheless, the microbiome
field is making progress to avoid some of these pitfalls including
validating their models on independent datasets (7, 18, 19) and
introducing analysis frameworks to better use ML tools (20--23). More
work is needed to further improve reproducibility and minimize
over-optimism for model performance.

Among microbiome researchers the lack of transparency in justifying a
modelling approach has been due to an implicit assumption that more
complex models are better because they are more complex. This has
resulted in a trend towards using models such as random forest and
neural networks (2, 11, 24--26) over simpler models such as logistic
regression or other linear models (18, 22, 27). Although the more
complex models may be better at incorporating non-linear relationships
or yield better predictions, they are considered to be black box models
because they are not inherently interpretable. These models require post
hoc explanations to quantify the importance of each feature in making a
prediction and they do not show the structure of how the features are
used. Depending on the application of the model, researchers may choose
to use different modeling approaches. For example, researchers trying to
identify the populations causing a disease would likely want a more
interpretable model whereas clinicians may emphasize performance.
Although one may feel that they are sacrificing interpretability for
performance, that tradeoff may be minimal (28, 29). Regardless, it is
important for researchers to articulate why they have selected a
specific modelling approach or even compare multiple approaches in the
same study.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing 7 modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also built four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared the
predictive performance, interpretability, and computational efficiency.
To demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classifiy individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (3). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (1, 3, 4, 30). With this dataset we developed a framework
that implementf a ML pipeline that can be used for any modeling
approach, evaluates predictive performance, and demonstrates how to
interpret these models. This framework can be easily applied to other
host-associated and environmental microbiome datasets.

@@

\textbf{Model selection and pipeline construction} We established a ML
pipeline where we trained and validated each of the seven models using a
common approach that is based on standard methods within the ML
community (REFS){[}Figure 1{]}.

@@

Second, we trained seven different models using the training data. We
selected models with different classification algorithms and
regularization methods. Regularization is a technique that discourages
overfitting by penalizing the model for learning the training data too
well. For regularized logistic regression and SVM with linear kernel, we
used L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data led to a sparser solution
(i.e., force many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including decision tree, random forest, and
XGBoost and we trained an SVM model with non-linear kernel.

Third, fitting of these models require selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance. To select the best performing hyperparameter, we
performed a full grid search for hyperparameter settings when training
our models. Default hyperparameter settings in previously developed ML
packages in R, Python, and Matlab programming languages are inadequate
for effective application of classification algorithms and need to be
optimized for each new dataset used to generate a model. In the example
of L1-regularized SVM with linear kernel {[}Figure S1{]}, the model
showed large variability between different regularization coefficients
(C) and was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularlized logistic regression and
linear SVM) performed nearly as well as random forest.

@@

To evaluate the risk for over-optimism of each model, we calculated the
range of AUROC values for each model using 100 splits. The range among
the testing AUROC values within each model varied by 0.23 on average
across the seven models. If we had only done a single split, then there
is a risk that we could gotten lucky or unlucky with the performance of
the model. For instance, the lowest AUROC value of the random forest
model was 0.593 whereas the highest was 0.81. These results showed that
depending on the data-split, the testing AUROC values showed great
variability {[}Figure 2{]}. Therefore, it is important to employ the
hierarchical data splits that were included in our pipeline to minimize
the risk of over-optimism.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
range among the cross-validation AUROC values within both models at
lower sample sizes were much larger than when the full collection of
samples was used to train and validate the models. These results showed
that because the microbiome data had many features (6920 OTUs), it was
important to train the models using appropriate sample sizes to avoid
problems with generalizability and over-optimism. Furthermore, it was
encouraging that even for a small number of samples, the 95\% confidence
interval included the median AUROC values for the larger subsetted
datasets.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed at the signs of feature weights to
determine whether an OTU is associated with classifying a subject as
being healthy or having an SRN - negative sign indicated being healthy
and positive sign indicated having SRN. Among the highest ranked OTUs
across these three models, it was encouraging that many of the same OTUs
(e.g.~OTU 50, 426, 609, 822, 1239) were important across all of the
models. The benefit of this approach is that the results of the analysis
are based on the trained model parameters and provide information
regarding the sign and magnitude of the impact of each OTU. However,
this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature
importance. Whereas the absolute feature weights were determined from
the trained models, here we measured importance using the held-out test
data. Permutation importance analysis is a posthoc explanation of the
model where we randomly permuted non-correlated features individually
and groups of perfectly correlated features across the two groups in the
held-out test data. We then calculated how much the predictive
performance of the model (i.e testing AUROC values) decreased when each
OTU or group of OTUs was randomly permuted. We ranked the OTUs based on
how much the median testing AUROC decreased when it was permuted; the
OTU with the largest decrease ranked highest {[}Figure 4{]}. Among the
twenty OTUs with the largest impact for each of these models, there was
only one OTU (OTU 822) that was shared among all of the models; however,
we found three OTUs (OTU 58, 110, 367) that were important in each of
the tree-based models. Similarly, the random forest and XGBoost models,
shared four of the most important OTUs (OTU 2, 12, 361, 477).
Permutation analysis results also revealed that with the exception of
the decision tree model, removal of any individual OTU had a minimal
impact on model performance. For example, if OTU 367 were permuted
across the diagnoses from the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if we permuted the same OTU
from the random forest model, the AUROC only dropped from 0.695 to
0.6799. Effectively, the complexity of the communities was more fully
represented in the better performing models(22, 32). At least in this
case, it was not possible to distinguish between health and disease
using a single OTU. While permutation analysis allowed us to gauge the
importance of an OTU, the analysis is post-hoc (i.e.~done using the test
data) and these results did not allow us to directly interrogate the
models to know whether an OTU is associated with classifying a subject
as being healthy or having an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S3{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S3{]}, 17 out of the 20 top OTUs
(e.g.~OTU 609, 822, 1239) were deemed important by both approaches.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
approaches. These results indicate that both approaches are consistent
in selecting the most important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. As expected, the training times
increased with the complexity of the model and the number of tuned
hyperparameter settings. Also, the linear models trained faster than
non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
size of the training dataset, we observed a {[}linear, quadratic, etc{]}
relationship between the size of the dataset and the training time for
the random forest model.

@@

Overview • Establish pipeline that can be generalized to any modeling
method • Linear/non-linear not a bfd • Interpretability tradeoff Point 1
• Importance for transparency, picking the most complex model is not
necessarilty the best model • Good / better / best: Table S1 • ``best''
has multiple meanings Point 2 • Our paper is concerned with developing
models to gain greater biological insights into what's going on • Still
need biological validation and further experimentation

Point 3 • These models show great opportunity to use microbiome to make
diagnoses • We talked about classification, but others have used it in
microbiome data to do regression, mulit-class -- pipeline is the same •
Others have used this on shotgun metagenomic data and collinearity is
expected to be an even larger problem since 5000 genes are all
correlated with each other • Deployment is a different question that
requires independent datasets to test on; we currently lack decent sized
parallel datasets where the same type of sequence data and clinical data
have been collected • We also have a problem with OTU-based approaches
Conclusion/Future direction • ML is awesome and there's always new
methods; our framework gives structure to investigators wanting to try
these out. We encourage reseraches to apply multiple models to their
data and report the results • Things like neural nets are attractive and
hold great promise, but we need a ton more data to apply and they are
impossible to interpret

There is a growing awareness that many human diseases and environmental
processes are not driven by single organism, but are the product of
multiple bacterial populations. Whereas traditional statistical
approaches are useful for identifying those cases where a single
organism can drive a process, ML methods offer the ability to use many
features of these communities to identify those cases. If it is possible
to classify communities reliably, then it is hoped that we could
identify those populations that are responsible for the classification.
Because application of ML in microbiome studies is still in its infancy,
we developed a framework to train rigorous, transparent and reproducible
ML models {[}Table 1{]}.

To test this framework, we benchmarked seven ML models to show that a
clearly defined ML problem that is based on the goal of the microbiome
study should inform our model selection. Our results showed that if the
goal of a study is to learn the ecology behind a disease and to identify
microbial biomarkers, we can create ML models that are inherently
interpretable and easily trained without losing predictive power. In
terms of predictive performance, random forest model had the best
testing AUROC values compared to the other six models. However, the
second best model was L2-regularized logistic regression with a median
AUROC difference of only 0.015 compared to random forest. While random
forest took 83.2 hours to train, L2-regularized logistic regression
trained in 12 minutes. In terms of interpretability, random forest was a
more complex ML model and could only be explained using post-hoc methods
such as permutation importance. On the other hand, L2-regularized
logistic regression was easier to interpret by ranking absolute feature
weights of the trained model.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective model selection and implementation when applying machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective model selection and implementation when applying
machine learning to bacterial 16S rRNA gene sequencing data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Mack
Ruffin\({^3}\), Jenna Wiens\({^2}\), Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify the microbial biomarkers and aid in diagnosis of many
chronic diseases such as inflammatory bowel disease, diabetes, and
colorectal cancer. Progress has been made towards developing ML models
that predict health outcomes from bacterial abundances, but inconsistent
adoption of methods call the validity of these models into question.
Furthermore, there appears to be a preference by many researchers to
favor increased model complexity over interpretability. To overcome
these challenges, we trained seven models that used fecal 16S rRNA
sequences to predict the presence of colonic screen relevant neoplasias
(SRNs; n=490 patients, 261 controls and 229 cases). We developed a
generalizable pipeline to train, validate and interpret the models. To
show the effect of model selection, we assessed the predictive
performance, interpretability, and computational efficiency of the
following models: L2-regularized logistic regression, L1 and
L2-regularized support vector machines (SVM) with linear and radial
basis function kernels, a decision tree, random forest, and extreme
gradient boosting (XGBoost). The random forest model was best at
detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h)
and hard to interpret. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680, it trained much faster (12 min) and was more
interpretable. Our analysis showed that ML models should be chosen based
on the goal of the study as it will inform our expectations of
predictive performance and interpretability.

@@

\subsection{Importance}\label{importance}

Prediction of health outcomes using machine learning (ML) is rapidly
being adopted in human microbiome studies. However, these ML models are
likely overoptimistic in terms of quantifying predictive performance.
Moreover, there is a trend towards using black box models such as random
forest and neural networks without a discussion of the difficulty of
interpreting such models when trying to identify microbial biomarkers of
disease. This work represents a step towards developing better ML
practices in microbiome research by implementing a rigorous pipeline and
emphasizing the importance of selecting ML models that reflect the goal
of the study. These concepts are not particular to study of health
outcomes but can also be applied to environmental microbiology studies
as well.

@@

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual with high confidence
relies on a model that is built with rigorous methods. However, there
are common methodological problems across many of these studies that
need to be addressed as the field progresses. These include a lack of
transparency in which methods are used and how these methods are
implemented; developing and evaluating models without a separate
held-out test data; large variation between the predictive performance
on different folds of cross-validation; and large variation between
cross-validation and testing performances. Nevertheless, the microbiome
field is making progress to avoid some of these pitfalls including
validating their models on independent datasets (7, 18, 19) and
introducing analysis frameworks to better use ML tools (20--23). More
work is needed to further improve reproducibility and minimize
over-optimism for model performance.

Among microbiome researchers, the lack of transparency in justifying a
modelling approach has been due to an implicit assumption that more
complex models are better because they are more complex. This has
resulted in a trend towards using models such as random forest and
neural networks (2, 11, 24--26) over simpler models such as logistic
regression or other linear models (18, 22, 27). Although the more
complex models may be better at incorporating non-linear relationships
or yield better predictions, they are considered to be black box models
because they are not inherently interpretable. These models require post
hoc explanations to quantify the importance of each feature in making a
prediction and they do not show the structure of how the features are
used. Depending on the application of the model, researchers may choose
to use different modeling approaches. For example, researchers trying to
identify the populations causing a disease would likely want a more
interpretable model whereas clinicians may emphasize predictive
performance. Although one may feel that they are sacrificing
interpretability for performance, that tradeoff may be minimal (28, 29).
Regardless, it is important for researchers to articulate why they have
selected a specific modelling approach or even compare multiple
approaches in the same study.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing 7 modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also built four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared their
predictive performance, interpretability, and computational efficiency.
To demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classifiy individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (3). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (1, 3, 4, 30). With this dataset we developed a framework
that implements a ML pipeline that can be used for any modeling
approach, evaluates predictive performance, and demonstrates how to
interpret these models. This framework can be easily applied to other
host-associated and environmental microbiome datasets.

@@

\textbf{Model selection and pipeline construction} We established a ML
pipeline where we trained and validated each of the seven models using a
common approach that is based on standard methods within the ML
community (REFS by Jenna?){[}Figure 1{]}.

@@

Second, we trained seven different models using the training data
{[}Table 1{]}. We selected models with different classification
algorithms and regularization methods. Regularization is a technique
that discourages overfitting by penalizing the model for learning the
training data too well. For regularized logistic regression and SVM with
linear kernel, we used L2 regularization to keep all potentially
important features. For comparison, we also trained an L1-regularized
SVM model with linear kernel. L1-regularization on microbiome data led
to a sparser solution (i.e., force many coefficients to zero). To
explore the potential for non-linear relationships among features to
improve classification, we trained tree-based models including decision
tree, random forest, and XGBoost and we trained an SVM model with
non-linear kernel.

Third, fitting of these models require selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance. To select the best performing hyperparameter, we
performed a full grid search for hyperparameter settings when training
our models. Default hyperparameter settings in previously developed ML
packages in R, Python, and Matlab programming languages are inadequate
for effective application of classification algorithms and need to be
optimized for each new ML task. In the example of L1-regularized SVM
with linear kernel {[}Figure S1{]}, the model showed large variability
between different regularization coefficients (C) and was susceptible to
performing poorly if the wrong regularization coefficient was assigned
to the model by default.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and SVM
with linear kernel) performed nearly as well as random forest.

@@

To evaluate the risk for over-optimism of each model, we calculated the
range of AUROC values for each model using 100 data-splits. The range
among the testing AUROC values within each model varied by 0.23 on
average across the seven models. If we had only done a single split,
then there is a risk that we could gotten lucky or unlucky with the
performance of the model. For instance, the lowest AUROC value of the
random forest model was 0.593 whereas the highest was 0.81. These
results showed that depending on the data-split, the testing AUROC
values showed great variability {[}Figure 2{]}. Therefore, it is
important to employ the hierarchical data splits that were included in
our pipeline to minimize the risk of over-optimism.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
range among the cross-validation AUROC values within both models at
lower sample sizes were much larger than when the full collection of
samples was used to train and validate the models. These results showed
that because the microbiome data had many features (6920 OTUs), it was
important to train the models using appropriate sample sizes to avoid
problems with generalizability and over-optimism. Furthermore, it was
encouraging that even for a small number of samples, the interquartile
range included the median AUROC values for the larger subsetted
datasets.

@@

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU is associated with classifying a subject as
being healthy or having an SRN - negative sign indicated being healthy
and positive sign indicated having an SRN. It was encouraging that many
of the highest ranked OTUs were shared across these three models,
(e.g.~OTU 50, 426, 609, 822, 1239). The benefit of this approach was
that the results of the analysis were based on the trained model
parameters and provided information regarding the sign and magnitude of
the impact of each OTU. However, this approach was only possible with
linear models.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a posthoc explanation of the model where we randomly permuted
non-correlated features individually and groups of perfectly correlated
features across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e testing
AUROC values) decreased when each OTU or group of OTUs was randomly
permuted. We ranked the OTUs based on how much the median testing AUROC
decreased when it was permuted; the OTU with the largest decrease ranked
highest {[}Figure 4{]}. Among the twenty OTUs with the largest impact,
there was only one OTU (OTU 822) that was shared among all of the
models; however, we found three OTUs (OTU 58, 110, 367) that were
important in each of the tree-based models. Similarly, the random forest
and XGBoost models, shared four of the most important OTUs (OTU 2, 12,
361, 477). Permutation analysis results also revealed that with the
exception of the decision tree model, removal of any individual OTU had
a minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.68.
Effectively, the complexity of the communities was more fully
represented in the better performing models (22, 32). At least in this
case, it was not possible to distinguish between health and disease
using a single OTU. While permutation analysis allowed us to gauge the
importance of an OTU, the analysis was post-hoc (i.e.~done using the
test data) and these results did not allow us to directly interrogate
the models to know whether an OTU is associated with classifying a
subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S4{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S4{]}, 17 out of the 20 top OTUs
(e.g.~OTU 609, 822, 1239) were deemed important by both interpretation
methods. Similarly, for the L2-regularized SVM and L2-regularized
logistic regression, 9 and 12 OTUs, respectively, were shared among the
two interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. As expected, the training times
increased with the complexity of the model and the number of tuned
hyperparameter settings. Also, the linear models trained faster than
non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
size of the training dataset, we observed a linear relationship between
the size of the dataset and the training times for L2-regularized
logistic regression and random forest models {[}Figure S5{]}.

@@

There is a growing awareness that many human diseases and environmental
processes are not driven by a single organism but are the product of
multiple bacterial populations. Traditional statistical approaches are
useful for identifying those cases where a single organism is associated
with a process. In contrast, ML methods offer the ability to incorporate
the structure of the microbial communities as a whole to classify them
into different categories such as coming from a patient who is healthy
or has SRNs. If it is possible to classify communities reliably, then ML
methods also offer the ability to identify those microbial populations
within the communities that are responsible for the classification.
However, the application of ML in microbiome studies is still in its
infancy and the field still needs to develop a better understanding of
different ML methods, their strengths and weaknesses, and how to
implement them.

To address these needs, we developed a framework to train rigorous,
transparent, and reproducible models. We benchmarked seven ML models and
showed that we can create models that are inherently interpretable and
easily trained without losing predictive performance. In terms of
predictive performance, the random forest model had the best testing
AUROC values compared to the other six models. However, the second-best
model was L2-regularized logistic regression with a median AUROC
difference of only 0.015 compared to random forest. While random forest
took 83.2 hours to train, L2-regularized logistic regression trained in
12 minutes. In terms of interpretability, random forest was a complex ML
model and could only be explained using post-hoc methods such as
permutation importance. On the other hand, L2-regularized logistic
regression was easier to interpret by ranking the OTUs based on their
feature weights and reviewing the signs of these weights. Comparing many
different models showed us that the most complex model is not
necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that
can be generalized to any modeling method that uses 16S rRNA sequence
counts to predict a binary health outcome. We performed a random
datasplit to create a training set (80\% of the data) and a held-out
test set (20\% of the data), which we used to evaluate predictive
performance. We repeated this datasplit 100 times to measure predictive
performance. During the training, we tuned the model hyperparameters
with a repeated five-fold cross-validation. Despite the high number of
features microbiome datasets typically have, the models we built with
this pipeline were generalizable as shown by the similar AUROC values
from the cross-validation and testing.
